Spring haul-out behavior of seals in the Bering and Chukchi seas

### 2023-02-07 analysis/paper/reviewer_response.md

We have moved this reference to later in the paragraph so the first
references are about Arctic work. While the example is from the
Antarctic, it is an important, early example of analysis using
bio-logger data to estimate haul-out behavior.

> *Do you mean age-class? The term life-history implies knowledge of
> previous behavioral efforts or occupancy of regions of interest.  
> Here you are specifically interested in age-class.*

We have edited to use the term 'demographic'

> Please help your reader understand why are you doing your analysis and
> not offer a wide and jargon rich inventory of concepts that may be
> considered.

> Please keep your introduction focused on establishing the knowledge
> gap that you are seeking to fill. Please do not lead your reader to
> the numerous other knowledge gaps that are not being addressed by your
> current study.

@@

We have removed the sentence in question

> Please use the term "hauled-out". Neither the biologger nor the aerial
> survey is able to distinguish among the various seal behaviors that
> result in them being hauled out and available for view by an aerial
> survey.

@@

noted, and the terms 'hauled out', 'to haul out', etc is used
consistently throughout the manuscript

> You have indicate that you would consider sea ice extent. Here you
> introduce the concept of icescape. Please introduce this concept
> earlier and indicate how you will incorporate this into your analysis.

@@

> Please use an active voice for clarity.

This sentence has been changed to:

"For this study we used haul-out behavior data and location estimates
from bio-loggers deployed on bearded, ribbon, and spotted seals in the
Bering, Chukchi, and western Beaufort seas by multiple organizations as
part of collaborative investigations from 2005 through 2020."

> You summarize the years during which land-based studies applied
> bio-loggers.\
> Please do the same for the offshore deployments.

@@

This sentence now reads:

"Ship-based capture events occurred during spring near the southern ice
edge in the Bering Sea between 2005 and 2018."

> Please provide the geographic context of your tag deployments and data
> collection using a plot similar to that used by Jay et al. (2012:
> 10.3354/meps10057 figure 2). Please consider using a tabulation that
> enables readers to understand the geographic context of the
> deployments similar to that used by Jay et al. 2012, table 1.

@@

We feel the spatial distribution of the actual transmitted data --
compared to the deployment location -- is most informative and relevant
for the audience's understanding of how our data are distributed
spatially and how well data represents the range of each species. We
provide this visually in Figure 3 and adding an additional layer to that
figure would to show deployments would be an additional complication. To
ensure the audience as a clear understanding of our study area, we have
added an additional figure with a map of the Bering and Chukchi sea
region that indicates generally where deployments started. We have also
added a more detailed figure to the supplemental section with each
location specifically marked.

> Please use active voice for all actions that you took in your data
> collection, preparation and analysis.

@@

Noted and addressed

> Please move this to the discussion. This is a summary of your data
> findings, not a description of your methods.

@@

We can see the reviewer's point. However, we do not feel this topic is a
summary of our data findings in this paper. Instead, our view is that
this information provides the context and rationale for our
methodological choice regarding bio-logger type and attachment methods.
It is also important regarding the analytic methods -- preference for
flipper tag haul-out data over those from a tag adhered to the
dorsal/head. If we were to move these important points to the
discussion, we worry those points would be left disconnected.

> This is a discussion of your methods. This is not your methods. Please
> state how the data was collected in your Methods section and discuss
> how data collection have been uncertain or biased in your discussion
> section.

@@

We appreciate this suggestion from the reviewer and we have moved large
portions of this paragraph to the discussion. However, some information
regarding age/sex class determination is left here so the reader has
context for Table 1.

> Please indicate how this was determined during data collection.

We have edited this sentence some to clearly communicate that field
determination of reproductive status (and, thus ideal delineation of
sub-adult and adult age class) was not possible.

> Please allow your readers the opportunity to understand the geographic
> context of these deployments through use of symbols that maybe used to
> cross-reference the deployment locations in a map that indicates
> deployment locations along with filtered movement tracks.

@@

As previously addressed, there are now two figures available that
provide geographic context of the deployments. Figure 3 shows the
geographic distribution of the data from the deployments. The paper is
not presenting or discussing movement ecology and, therefore, we do not
present filtered movement tracks. We keep the focus on haul-out behavior
data.

> Please consider bar plot similar to that used by Jay et al. (2012:
> 10.3354/meps10057 figure 3), replacing the region with the seal
> species.

@@

This figure has been updated with this suggestion in mind

> Please state clearly how you determined the end of a deployment. Here
> you indicate various elements of the haulout chronology, movement and
> dive chronology that may have been used, however not reader could
> replicate your methods base don this text.

@@

We've added the following text to the paragraph:

"For example, a data record that ends with several consecutive days
(\~10+ days) of 100% dry observations and with locations indicating the
tag was on land would be truncated to not include the final stretch of
100% dry observations. The vast majority of deployments end with the
device detaching in the water and the deployment end date is obvious.
There is no perfect algorithm for identifying deployment end dates and
each deployment in question must be considered separately. While not
perfect, we are confident our reliance on expert opinion and examination
of multiple data streams provides the best option."

> Deployment activity is known to disrupt normal pinniped behavior. Did
> you censure some period of data collection during and after the
> deployment to ensure that that disruptions from the deployment
> activity did not bias your analysis?

@@

For these phocid species, we have no empirical data (or even anecdotal
observational data) on how captures may or may not disrupt 'normal'
behavior. This is likely to be influenced by the capture method, how
long the animal was restrained, whether the animal was with a dependent
pup, whether the animal was molting, environmental variables such as
wind and temperature, etc etc. Given all these unknowns, we have not
applied a censure to the beginning of our deployment data. We can't 100%
ensure this choice doesn't result in some level of bias. However, given
the large sample size and duration of deployments we expect any bias
imposed is negligible.

> This is a result. Please indicate the method that you used to
> distinguish "plausible" from "implausible" telemetered seal behavior.
> In your Results section, you may indicate how method was able to
> identify occasions of implausible seal behavior and any patterns that
> arose relative to tag model or attachment.

This sentence has been moved to the results section. Additionally, we've
provided a brief explanation of implausible along with a reference for
more details. For these devices, the percent-dry timeline reported 100%
dry for several weeks followed by an additional period of time that was
more typical of expected bearded seal behavior. During the extended dry
time the locations indicated movement of the tag that would not be
simply drifting ice. We also observed increasing transmission rates. The
working (i.e. unconfirmed guess) is that the wet/dry sensor surface
became obscured by some corrosive element such that when in the water
conductivity was prevented and, thus, recorded as dry.

> Please move to the Results section. Please move the discussion of how
> molt timing affects termination of tag deployments to the discussion
> section.

This figure has been moved to the results section along with additional
descriptive text. We have left the discussion regarding molt timing and
termination of the data record here to better communicate the point to
the reader and prevent an isolated mention in the discussion that has no
anchor.

> Thus, the a successful transmission of data was required at no greater
> than 48 h intervals.

We're not sure how to interpret this comment. The main point of the
sentence is to explain the rationale/benefits of transmitting the
limited selection of percent dry values. We have expanded the text to
provide more details regarding the caching of messages and data
transmission.

> Do you mean "atmospheric" magnetic interference.\
> Presumably "extra-terrestrial atmospheric" events (such as events on
> Venus or Mars) played no role in the transmission of the data signal
> to the Argos receivers in orbit around Earth.

We were using a less strict definition of extra-terrestrial to refer to
interference that wasn't Earth/surface-based. We've edited to
'atmospheric interference'

> This note belongs in the Discussion, not the Results.

We feel this note is important to acknowledge in the methods because it
speaks to the rationale for our methodological choices (bio-logger type,
programming choices, analytical choices).

> Do you mean that each cell "encompasses" and area of 50 km\^2?

We have edited to the reviewer's suggested word 'encompasses'

> Please use the active voice.

Thank you for catching this. We have edited this sentence to active
voice

> For most tags, your daily locations are from the Argos system, not
> from your data loggers. However, you included some FastLoc tag
> deployments. Please re-write for clarity of what these location are.

The previous paragraph describes how we determined daily locations. We
see the reviewers point re: the data not technicaly coming from our data
loggers. We have edited the sentence for additional clarity.

> What is a "key" weather variable? Please write for clarity.

We've edited this sentence removed the word 'key'

> This is a procedural detail. Other readers may replicate your method
> without needing to do this. Please omit this detail.

> Readers can replicate your methods without the details on converting
> the NARR wind vectors into wind speed.

We're not sure why the reviewer would recommend omitting a 'procedural
detail' as part of our methods. We feel it is important to clearly
communicated our procedure for processing the large NARR data set and
that we used the native grid size of 32km. Regarding the calculation of
wind vectors, we include this detail because many readers, like us
initially, may not be well versed in how wind vectors components are
converted to wind speed.

> Please use an active voice: "For all seal species we considered the
> following variables when modeling the hourly haulout behavior..."

We have edited this sentence with the reviewer's suggestion

> Using the term "major focus" implies that there may be numerous minor
> foci. Can you not state that this is the goal of the study? Ancillary
> natural history insights can be discussed incidental to this, but are
> not driving the research. The statutory requirements of the MMPA and
> ESA are driving your study. These Federal laws require robust methods
> to monitor population status. Your study seeks to build this method.
> Discussion of natural history belong in your Discussion section and
> not in the framing of your study. Please help the reader focus on why
> the Federal government is doing this study.

> Understanding pinniped haulout behavior is essential to designing and
> interpreting population estimation aerial surveys. Understanding how
> they use sea ice during a time when sea ice availability is rapidly
> changing is also important, but is ancillary to the paper's focus.

We have removed the term 'major focus' and, as the reviewer suggests,
simply state that this is the goal of the study. We have also added
mention of the statutory requirements to paragraph 2 of the Introduction
section and expanded the text to acknowledge the importance of
understanding how seals use sea ice during this period of change but,
ultimately, it's ancillary to the focus of this paper.

As for providing the reader focus on why the Federal government is doing
this study: this is a collaborative study across multiple organizations
and the analysis/purpose is not limited to the Federal government. We've
incorporated the reviewers suggests to expand discussion of the
statutory requirements of the MMPA and ESA but don't feel we need to
single out the motivations for the Federal government.

> Seals can just as happily be on foraging trips within the pack and
> marginal sea ice as in the open water, depending on the species.
> Suggesting that foraging trips are associated with ice-free conditions
> is misleading.

We have removed 'e.g., foraging trips' from this sentence so as not to
mislead the reader.

> This is your primary reason for not including sea ice geography in
> your analysis: you simply lacked geolocation of much of your haulout
> behavior data.
>
> Please help the reader understand how coarse your tracking data was
> for the flipper tagged animals and what proportion of each species was
> tracked with these flipper tags. Please also help the reader
> understand how the tracking data relates geographically to regions
> that may be subject to aerial survey efforts (i.e. are your tracking
> data collected from across the regions that may be surveyed during
> range-wide spring aerial surveys?).

We have re-organized this paragraph to, hopefully, better communicate our
choice not to explore sea-ice related covariates. As the reviewer suggests,
we have limited geolocation data for at-sea periods of our haul-out behavior
data -- we know the seal was in the water but don't have any estimates of
location when only a flipper tag is attached. By leading with this statement,
we hope that communicates our primary rationale. But, the other rationales
are also important and we still include them.

We have also provided some statistics regarding the percentage of the
percent-dry timeline data that is not associated with a corresponding location
estimate because the seal was in the water and no estimates were available from
flipper mounted SPOT tags.

> Please help the reader understand why you have not included a basic
> metric of sea ice geography, such as distance from marginal ice edge
> (from the NIC's daily Marginal Ice Zone charts), which is a metric
> that would be available for all data collected during an aerial
> survey.

> It is a bit misleading to indicate that the NARR variables were
> sourced from NARR alone. Their values were interpolated onto a
> geogrpahic location that was widely variable in precision and that
> precision relates to the both the tag attachment (glued on or flipper
> attached) and the seal behavior (in water, hauled out).
>
> Indicating that your "Northing" variable was from the "Bio-logger" is
> also misleading because your locations was determined by either the
> Doppler inflection point time, satellite orbitography and a model of
> the Earth (the Argos relocation method) or by reprocessing of GNSS
> signals that were logged and then transmitted via the Argos system as
> a data message. Really, neither of these were determined by the
> "Bio-logger", rather they were determined by a very complex system of
> geolocation.

> Your readers may be better served by not including this "Source"
> field, given that you describe how you obtained or estimated these
> variable values in your methods text. Offering an over-simplified
> "Source" in the table obscures your actual methods.

We appreciate the reviewer's thoughts on this and agree with that removing
the 'Source' column better serves the readers.

### 2023-02-07 analysis/paper/London_HauloutBeahvior.Rmd
more edits

Statutory requirements (e.g., United States Marine Mammal Protection Act (MMPA),
United States Endangered Species Act (ESA)) for timely estimates of population
abundance and trend of these species mean improved aerial survey effort is
needed -- and, must be paired with improved knowledge of haul-out behavior to
ensure accurate estimates. Several studies have contributed estimates of the
distribution and abundance of ice-associated seal species in the Arctic using
aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]). Such
abundance studies were conducted over very large areas and estimation of
absolute abundance required making inference about numerous issues affecting the
observation of seals on ice. These included availability (only seals on ice were
available to be counted), detection probability (observers or automated
detection systems may have missed some seals on ice), species misclassification,
and possible disturbance of seals by aircraft [@conn2014a; @verhoef2014a].
Refining these inferences will improve the accuracy of abundance estimates and,
hopefully, allow credible predictions about the effects of climate disruptions
on the abundance and distribution of Arctic seal populations.

@@

We subset haul-out behavior data from 
`r species_n %>% .$n %>% sum()` bio-loggers deployed on 
`r species_n %>% filter(species == "Bearded seal") %>% .$n` bearded, 
`r species_n %>% filter(species == "Ribbon seal") %>% .$n` ribbon, and 
`r species_n %>% filter(species == "Spotted seal") %>% .$n` spotted seals 
to include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and we either adhered them to the hair on the seal or attached
them through the rear flipper inter-digital webbing. The use of bio-loggers
adhered to the back or head provides some benefits over flipper mounted devices
(e.g. increased satellite transmittal rates, locations at sea) but these are
lost during the following annual molt, which, depending on deployment date,
limits the duration of haul-out data they provide. Additionally, bio-loggers
attached to the head or dorsal region are often dry while the seal is floating
at the surface, inducing a slight positive bias in the hourly percent-dry values
reported by the bio-logger. For this study, in cases where both bio-logger types
were deployed, we preferred hourly percent-dry observations from the flipper tag.>

@@

Haul-out behavior data were recorded and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and the percent of the
hour in the dry state was calculated (Figure \@ref(fig:examplePlot)). On board the
bio-logger, hourly percent-dry data were rounded to the nearest 10%
inclusive of 0% and 100% and additional values at 3% and 98%. This
compression resulted in additional data transmission as each message consisted
of two complete 24-hour records. Bio-loggers were, typically, deployed and
programmed in a manner to maximize data transmission during the spring pupping
and molting period, though hourly percent-dry data were not always
successfully transmitted. This is due to a variety of factors including
satellite coverage, tag availability (e.g., tags mounted to the rear flipper
often do not transmit while at sea), tag performance, duty cycling, and
extra-terrestrial atmospheric interference. Fortunately, missing records do not
substantially bias inference about haul-out probabilities [@conn2012b].

@@

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and seal location. We explored the use
of a continuous-time correlated random walk [@johnson2008a] movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_ = 250 m, _2_ = 500 m, _1_ = 1500 m, _0_ = 2500 m [@lopez2013]; we chose 2500 m for
location classes _A_ and _B_) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50 m. On days when haul-out observations were
present but location data were missing we used the seal's last calculated weighted
average daily location; days when the location intersected with land
were removed from the seal's record. We recognized that bearded and spotted
seals haul out on land. However, assessing the relationship between haul-out
behavior and weather covariates and seals' availability for aerial surveys on
land was outside the scope of this study. Additionally, any daily locations on
land were likely more reflective of coordinate averaging and measurement error,
rather than actual use of coastal features. Figure \@ref(fig:dataMap) shows the
spatial distribution of weighted locations with available haul-out behavior data
used for analysis of each species across the study area.

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). Solar hour was calculated
using the {solaR} package [@perpinan2012] within the R statistical environment
[@rcoreteam2021] using the daily locations. We also linked the weighted average
daily locations from our bio-loggers to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made
available across the region 8 times daily. For this study, NARR weather values
were subset to the extent of our study area over the Bering and Chukchi seas at
3-hr intervals based on the native grid resolution of 32 km (1024 km^2^). The
following meteorological variables are known to affect haul-out behavior in
other Arctic pinnipeds [@reder2003;
@udevitz2009; @perry2017] and were interpolated and assigned to daily seal
locations using a bilinear method: 1) air temperature at 2 m above the Earth's
surface, 2) wind consisting of northerly and easterly vector components
converted to wind speed using the Euclidean norm, 3) barometric pressure at sea
level, and 4) precipitation (Table \@ref(tab:tblCovariates)).

Models for all species included the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill
(represented by a *wind:temperature* interaction [@udevitz2009]), and
day-of-year and time-of-day (solar hour) interactions to permit diurnal patterns
to change from March to July. Ribbon and spotted seal models included age-sex
class and interactions between day-of-year and age-sex class, but we omitted
these from bearded seal models due to poor representation of age-sex classes
(Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a substantial
range during the spring and we were interested in possible differences in the
timing of haul-out behavior along a latitudinal gradient. We omitted the
latitudinal effect from ribbon and spotted seal models because, during the
spring, these species are most prevalent near the southern ice edge in the
Bering Sea [@conn2014a].

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem
counterintuitive when modeling the haul-out behavior of seal species with such a
close association to sea ice; seals haul out in the presence of sea ice, and we
could assess the local concentration of sea ice during these events (see
[@olnes2020]). However, a major focus of this study was to develop models
applicable for aerial survey correction factors and using sea ice as a covariate
would almost certainly bias haul-out predictions towards those seals that are on
or near ice and therefore more likely to haul out. Since aerial surveys can only
detect seals on ice, abundance estimates would be missing a correction for those
seals that are away from ice (e.g., on foraging trips). Additionally, many of
the deployments consisted of a single device attached to the rear flipper of the
seal which did not provide at-sea locations, limiting our ability to fully
evaluate fine-scale habitat preferences related to sea-ice. Lastly, our study
was limited to the spring season when seal haul-out tendencies are strongly
influenced by pupping, nursing, breeding behavior, and molt and these drivers
are likely more influential than specific sea-ice concentration. Crawford et al.
[-@crawford2019] compared haul-out probability models for ringed seals and found
those that only included season (and not sea-ice concentration) were the most
parsimonious. For these reasons, we have elected not to use sea ice
concentration _as a predictor for haul-out probability_ in the present study.

@@

covariate_tbl %>% 
  flextable() %>% 
  fontsize(size = 10, part = "all") %>% 
  hline(i=8) %>% 
  width(width = 1.0) %>% 
  width(j = ~ Description, width = 2.5)
```

>>>>

Statutory requirements (e.g., United States Marine Mammal Protection Act (MMPA),
United States Endangered Species Act (ESA)) for timely estimates of population
abundance and trend of these species mean improved aerial survey effort is
needed -- and, must be paired with improved knowledge of haul-out behavior to
ensure appropriate survey design, robust methods, and accurate estimates.
Several studies have contributed estimates of the distribution and abundance of
ice-associated seal species in the Arctic using aerial surveys (e.g.,
[@bengtson2005a], [@conn2014a], and [@verhoef2014a]). Such abundance studies
were conducted over very large areas and estimation of absolute abundance
required making inference about numerous issues affecting the observation of
seals on ice. These included availability (only seals on ice were available to
be counted), detection probability (observers or automated detection systems may
have missed some seals on ice), species misclassification, and possible
disturbance of seals by aircraft [@conn2014a; @verhoef2014a]. Refining these
inferences will improve the accuracy of abundance estimates and, hopefully,
allow credible predictions about the effects of climate disruptions on the
abundance and distribution of Arctic seal populations.

@@

We subset haul-out behavior data from 
`r species_n %>% .$n %>% sum()` bio-loggers deployed on 
`r species_n %>% filter(species == "Bearded seal") %>% .$n` bearded, 
`r species_n %>% filter(species == "Ribbon seal") %>% .$n` ribbon, and 
`r species_n %>% filter(species == "Spotted seal") %>% .$n` spotted seals 
to include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and we either adhered them to the hair on the seal or attached
them through the rear flipper inter-digital webbing. The use of bio-loggers
adhered to the back or head provides some benefits over flipper mounted devices
(e.g. increased satellite transmittal rates, locations at sea) but these are
lost during the following annual molt, which, depending on deployment date,
limits the duration of haul-out data they provide. Additionally, bio-loggers
attached to the head or dorsal region are often dry while the seal is floating
at the surface, inducing a slight positive bias in the hourly percent-dry values
reported by the bio-logger. For this study, in cases where both bio-logger types
were deployed, we preferred hourly percent-dry observations from the flipper tag.

@@

Haul-out behavior data were recorded and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and the percent of the
hour in the dry state was calculated (Figure \@ref(fig:examplePlot)). On board
the bio-logger, hourly percent-dry calculations were rounded to the nearest 10%
inclusive of 0% and 100% along with additional values at 3% and 98%. This
compression resulted in additional data transmission as each message consisted
of two complete 24-hour records. Memory capacity allowed caching of percent-dry
records for several weeks or months and each message was transmitted several
times to ensure reception at the satellite. Bio-loggers were, typically,
deployed and programmed in a manner to maximize data transmission during the
spring pupping and molting period, though hourly percent-dry data were not
always successfully transmitted. This is due to a variety of factors including
satellite coverage, tag availability (e.g., tags mounted to the rear flipper
often do not transmit while at sea), tag performance, duty cycling, and
atmospheric interference. Fortunately, missing records do not substantially bias
inference about haul-out probabilities [@conn2012b].

@@

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and seal location. We explored the use
of a continuous-time correlated random walk [@johnson2008a] movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_ = 250 m, _2_ = 500 m, _1_ = 1500 m, _0_ = 2500 m [@lopez2013]; we chose 2500 m for
location classes _A_ and _B_) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50 m. On days when haul-out observations were
present but location data were missing we used the seal's last calculated weighted
average daily location; days when the location intersected with land
were removed from the seal's record. We recognize that bearded and spotted
seals haul out on land. However, assessing the relationship between haul-out
behavior and weather covariates and seals' availability for aerial surveys on
land was outside the scope of this study. Additionally, any daily locations on
land were likely more reflective of coordinate averaging and measurement error,
rather than actual use of coastal features. Figure \@ref(fig:dataMap) shows the
spatial distribution of weighted locations with available haul-out behavior data
used for analysis of each species across the study area.

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). We calculated local solar
hour using the {solaR} package [@perpinan2012] within the R statistical
environment [@rcoreteam2021] based on the weighted daily average locations. We
also linked the weighted average daily locations to weather values from the
North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made available
across the region 8 times daily. For this study, NARR weather values were subset
to the extent of our study area over the Bering and Chukchi seas at 3-hr
intervals based on the native grid resolution of 32 km (1024 km^2^). The
following meteorological variables are known to affect haul-out behavior in
other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017] and were
interpolated and assigned to daily seal locations using a bilinear method: 1)
air temperature at 2 m above the Earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the
Euclidean norm, 3) barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

For all seal species, we considered the following 
variables when modeling the hourly haul-out behavior: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill
(represented by a *wind:temperature* interaction [@udevitz2009]), and
day-of-year and time-of-day (solar hour) interactions to permit diurnal patterns
to change from March to July. Ribbon and spotted seal models included age-sex
class and interactions between day-of-year and age-sex class, but we omitted
these from bearded seal models due to poor representation of age-sex classes
(Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a substantial
range during the spring and we were interested in possible differences in the
timing of haul-out behavior along a latitudinal gradient. We omitted the
latitudinal effect from ribbon and spotted seal models because, during the
spring, these species are most prevalent near the southern ice edge in the
Bering Sea [@conn2014a].

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem
counterintuitive when modeling the haul-out behavior of seal species with such a
close association to sea ice; seals haul out in the presence of sea ice, and we
could assess the local concentration of sea ice during these events (see
[@olnes2020]). This, however, expands the scope of our analysis into the realm
of habitat selection and many of our deployments consisted of a single device
attached to the rear flipper of the seal which did not provide at-sea locations,
limiting our ability to fully evaluate fine-scale habitat preferences related to
sea-ice. Insight into how seals use and interact with sea-ice during an extended
period when the availability and characteristics of sea-ice is rapidly changing
is important but ancillary to the focus of this analysis. The focus of this
study was to develop models applicable for aerial survey correction factors and
using sea ice as a covariate would almost certainly bias haul-out predictions
towards those seals that are on or near ice and therefore more likely to haul
out. Since aerial surveys can only detect seals on ice, abundance estimates
would be missing a correction for those seals that are away from ice. Lastly,
our study was limited to the spring season when seal haul-out behaviors are
strongly influenced by pupping, nursing, breeding behavior, and molt and these
drivers are likely more influential than specific sea-ice concentration.
Crawford et al. [-@crawford2019] compared haul-out probability models for ringed
seals and found those that only included season (and not sea-ice concentration)
were the most parsimonious. For these reasons, we have elected not to use sea
ice concentration _as a predictor for haul-out probability_ in the present
study.

@@

covariate_tbl %>% 
  select(-Source) %>% 
  flextable() %>% 
  fontsize(size = 10, part = "all") %>% 
  hline(i=8) %>% 
  width(width = 1.0) %>% 
  width(j = ~ Description, width = 2.5)
```
